Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2018-06-14 paper.tex
Make survival prob figures consistent with other figures
Change s->r in IV.B and simplify notation and wording

The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations for neutrinos
\cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths, i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions.
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.

But at the opposite limit, in the free-streaming regime,
coherence effects become important,
and the quantum kinetic equations take a Schr\"{o}dinger-like form:
\begin{equation}
  \label{eqn:schroedinger}
  i \frac{d}{ds} S = H S,
\end{equation}
where $ds$ is an interval of proper length traversed by the neutrino as
measured by our observer, and
$S$ is the neutrino flavor evolution matrix describing a mixed state
\begin{equation}
  \label{eq:density_evolution}
  \rho = S \, \rho_0 \, S^\dagger.
\end{equation}
In this limit, we may decompose the Hamiltonian into vacuum,
matter, and neutrino contributions:
\begin{equation}
  \label{eqn:decompose_H}
  H = H_{\rm V} + H_e + H_{\nu\nu}.
\end{equation}
Explicit formulas for these terms may be found in the oscillation literature,
e.g.\ \cite{duan2009-review},
but for clarity we give their order-of-magnitude scales here:
\begin{align}
  \label{eqn:scale_H_V}
  |H_{\rm V}| &\sim \frac{\Delta m^2}{\varepsilon}, \\
  \label{eqn:scale_H_e}
  |H_e|       &\sim G_{\rm F}\left|n_{e^-}-n_{e^+}\right|, \\
  \label{eqn:scale_H_nunu}
  |H_{\nu\nu}|&\sim G_{\rm F} \left|G_{\nu_e}-G_{\bar{\nu}_e}\right|,
\end{align}
where $\Delta m^2$ is the mass-squared differences between neutrino mass
eigenstates,
$G_{\rm F}$ is the Fermi coupling constant,
$n_{e^-}$ and $n_{e^+}$ are the electron and positron number densities,
and $G_{\nu_e}$ and $G_{\bar{\nu}_e}$ the electron neutrino and antineutrino
number densities defined in Eqn.~\ref{eqn:G}.
For the scale of $H_{\nu\nu}$ we have only included the isotropic
components of the neutrino fields.

In regimes in which $|H_{\nu\nu}|\ll|H_{\rm V},H_e|$,
the flavor evolution along the ray is soluble, and reveals a rich
and physically important phenonenology including
vacuum, solar, atmospheric, and terrestrial oscillations,
as well as oscillations in supernova envelopes.
Where neutrino densities are relatively high, however,
as in neutron star mergers, the problem must be solved globally.
To date, no method has been devised to handle this problem.

However, we can solve a similar but tractable problem along a single ray.
We assume that all the neutrino rays intersecting an event along a given test ray
have undergone the same flavor evolution history as that of the test ray,
i.e.\ $S$ is the same for all rays sharing that event.
This is the so-called single-angle approximation.
Though widely used in the supernova oscillation literature
and shown to be qualitatively faithful in those environments,
we note that recent studies have discovered configurations in which
the single-angle approximation masks certain instabilities
in the flavor evolution \cite{wu2017-fast_neutrino_conversions},
or introduces qualitatively different flavor evolution behavior
than that of the full multi-angle calculation \cite{vlas2018-multiangle}.

The formalism of the single-angle approximation requires knowledge of
the unoscillated neutrino contribution to the Hamiltonian
along a given test trajectory.
This is a function of the unoscillated neutrino self-interaction potential,
which for the $\alpha$-th flavor is:
\begin{equation}
  \label{eqn:V_nu_alpha}
  V_{\nu_\alpha,0}(\varepsilon,\ell_\mu) =
  \frac{\sqrt{2}G_{\rm F}}{(2\pi)^3} \varepsilon^2
  \oint d\Omega'(1-\varpi')f^{\nu_\alpha}(\varepsilon,\ell'_\gamma),
\end{equation}
with the test ray propagating in direction $\ell_\mu$,
ambient rays propagating in directions $\ell'_\gamma$,
and the cosine of the angle between these given by
$\varpi'=\psi^{\mu\gamma} \ell_\mu \ell'_\gamma$.
As in the moment equations (Eqns.~\ref{eqn:J}-\ref{eqn:Ka}),
the integral is taken over all directions $\ell'_\gamma$.

@@

If conditions are right, a resonant flavor transition
introducing significant mixing may occur very near
the point where neutrinos begin free-streaming.
The matter-neutrino resonance
\cite{malk2012-mnr_1,malk2015-mnr_2,malk2016-mnr_3},
can occur where the matter potential and
neutrino self-interaction potentials cancel,
or where $V_e - V_{\nu\nu,0}=0$, with
\begin{align}
  \label{eqn:V_e}
  V_e &= \sqrt{2} G_{\rm F} \frac{\varrho}{m_N} Y_e, \\
  \label{eqn:V_nunu0}
  V_{\nu\nu,0} &= \int d\varepsilon
  \big(V_{\nu_e,0}(\varepsilon)-V_{\bar{\nu}_e,0}(\varepsilon)\big),
\end{align}
with $\varrho$ the rest density and $m_{N}$ the nucleon mass.
The matter potential $V_e$ is always positive;
and far outside the accretion disks formed in neutron star mergers,
in which the disrupted neutron-star matter is rapidly releptonizing,
the total unoscillated neutrino self-interaction potential $V_{\nu\nu,0}$
is large and negative.

We examine this effect in the post-merger configuration already analyzed
in Sec.~\ref{ssec:test_disk_comparison}.
We calculate the self-interaction potential along a radial coordinate trajectory
making an angle $\theta=25^\circ$ with the rotation axis,
at $N_s=7$ positions $r\in\{30,50,82,135,223,368,608\}$~km.
We sample distribution functions at each of these positions over a grid
with extents $N_\varepsilon=30$, $N_A=200$, $N_B=30$;
energies range over $\varepsilon\in(0,100)$~MeV;
polar angles range over $\cos A \in ((\cos A)_{\rm min}, 1)$, with
$(\cos A)_{\rm min}\in\{-1,-1,-1,0.488,0.849,0.947,0.981\}$
for each of the $N_s$ positions;
and azimuthal angles range over the whole sky $B\in[0,2\pi)$.
We interpolate the logarithm of the self-interaction potentials
(Eqn.~\ref{eqn:V_nu_alpha})
in path length $\log s$ along the ray,
by fitting a 3rd order spline with continuous derivative,
across the $N_s$ observation points.
For $r>608$~km we extrapolate the self-interaction potentials using the
geometric fall-off of $r^{-4}$ applicable to the far-field limit of the
self-interaction potential \cite{malk2016-mnr_3}.

@@

In fact a spherical steady-state wind model, though providing an adequate
backdrop for the qualitative study presented in this section,
is less than ideal for this post-merger configuration,
most obviously because in the 10~ms since merger,
ejecta with the greatest velocities around $0.3c$
will have reached no further than $r_{\rm max} \sim 10^8$~cm.
Additionally, the true radial profile of the ejecta from this merger
(which our computational model does not follow)
will have many more features inside this radius, including shock jumps.
This model density profile is adequate as a backdrop to our study here, however,
since we expect the remnant to present similar neutrino emission over a thermal
timescale of a few tens of milliseconds
while the matter field propagates out to larger radii.

@@

We also solve for the flavor evolution of this system,
integrating $S$ according to Eqn.~\ref{eqn:schroedinger},
as described in \cite{zhu2016-mnr_nsns_remnant}.
We show the survival probabilities for $\nu_e$ and $\bar{\nu}_e$
for the \emph{noscat} treatment in Fig.~\ref{fig:survival-noscat},
and for the \emph{scat} treatment in Fig.~\ref{fig:survival-scat}.
The survival probability is the probability that a neutrino, if measured,
will be found to be in its original flavor state.
The survival probabilities are computed at each point along the trajectory
from the flavor evolution matrix,
$P_{\nu_\alpha \to \nu_\alpha} = |S_{\alpha \alpha}(E)|^2$.
When $P_{\nu_e \to \nu_e}$ decreases,
as can be seen for example in Fig.~\ref{fig:survival-scat} for $s>400$~km,
some of the electron-type neutrinos have oscillated into
mu- or tau-type neutrinos.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-survival_prob-noscat_ih}
  % Flavor evolution by Yonglin, see emails 5.9.18, 5.16.18
  % Data: $DIR/300-1499kmACC12Step2NE30IH051418m12sc.dat2p
  % Script: $DIR/plot.gp
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170412-self_interaction_potentials_M12_M12-LS220-M1/Run03/RunBA/RunBA-025deg-noscat/flavor_evolution
  \caption{Neutrino survival probabilities for the \emph{noscat} case.
    In the lower panel we reproduce the self-interaction potential $V_{\nu\nu,0}$
    and matter potential $V_e$ from Fig.~\ref{fig:V_nunu-noscat}.
    In this case, the neutrinos undergo collective neutrino oscillation
    beginning around $r\sim700$~km,
    with both the electron neutrinos and antineutrinos almost converting to mu
    or tau neutrinos completely.
  }
  \label{fig:survival-noscat}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-survival_prob-scat_ih}
  % Flavor evolution by Yonglin, see emails 5.9.18, 5.16.18
  % Data: $DIR/300-1499kmACC12Step2NE30IH051418m12sc.dat2p
  % Script: $DIR/plot.gp
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170412-self_interaction_potentials_M12_M12-LS220-M1/Run03/RunBA/RunBA-025deg-scat/flavor_evolution
  \caption{Same as Fig.~\ref{fig:survival-noscat},
    but for the \emph{scat} case.
    In this case the neutrinos undergo a standard MNR transition
    beginning around $r\sim400$~km,
    with the electron neutrinos converting to mu or tau neutrinos almost
    completely,
    and the electron antineutrinos oscillating back to their original flavor
    after partially converting to mu and tau antineutrinos.
  }
  \label{fig:survival-scat}
\end{figure}

@@

In addition to the fluid velocity and Lorentz factor, described above, the
other fluid state variables we use from our hydrodynamic simulations
are rest density $\varrho=m_b n_b$, temperature $T$, and electron fraction
\begin{equation}
  Y_e = \frac{n_{e^-} - n_{e^+}}{n_b},
\end{equation}
where $n_{e^-}$, $n_{e^+}$, and $n_b$ are the number densities of electrons
positrons, and baryons, and $m_b$ is the average baryon mass.

>>>>

The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations for neutrinos
\cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths, i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$ with
$H$ the Hamiltonian matrix describing coherent forward scattering interactions.
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.

But at the opposite limit, in the free-streaming regime,
coherence effects become important,
and the quantum kinetic equations take a Schr\"{o}dinger-like form:
\begin{equation}
  \label{eqn:schroedinger}
  i \frac{d}{ds} S = H S,
\end{equation}
where $ds$ is an interval of proper length traversed by the neutrino as
measured by our observer, and
$S$ is the neutrino flavor evolution matrix describing a mixed state
\begin{equation}
  \label{eq:density_evolution}
  \rho = S \, \rho_0 \, S^\dagger.
\end{equation}
In this limit, we may decompose the Hamiltonian matrix into vacuum,
matter, and neutrino contributions:
\begin{equation}
  \label{eqn:decompose_H}
  H = H_{\rm V} + H_e + H_{\nu\nu}.
\end{equation}
Explicit formulas for these matrices may be found in the oscillation literature,
e.g.\ \cite{duan2009-review},
but for clarity we only give their order-of-magnitude scales here:
\begin{align}
  \label{eqn:scale_H_V}
  |H_{\rm V}| &\sim \frac{\Delta m^2}{\varepsilon}, \\
  \label{eqn:scale_H_e}
  |H_e|       &\sim G_{\rm F}\left|n_{e^-}-n_{e^+}\right|, \\
  \label{eqn:scale_H_nunu}
  |H_{\nu\nu}|&\sim G_{\rm F} \left|G_{\nu_e}-G_{\bar{\nu}_e}\right|,
\end{align}
where $\Delta m^2$ is the mass-squared differences between neutrino mass
eigenstates,
$G_{\rm F}$ is the Fermi coupling constant,
$n_{e^-}$ and $n_{e^+}$ are the electron and positron number densities,
and $G_{\nu_e}$ and $G_{\bar{\nu}_e}$ the electron neutrino and antineutrino
number densities defined in Eqn.~\ref{eqn:G}.
In this statement of scale for $H_{\nu\nu}$ we only include the isotropic
components of the neutrino fields; for our calculations below, however, we
include the full angular distributions found via ray tracing.

In regimes in which $|H_{\nu\nu}|\ll|H_{\rm V},H_e|$,
the flavor evolution is locally soluble in a ray-by-ray method,
and reveals a rich and physically important phenonenology including
vacuum, solar, atmospheric, and terrestrial oscillations,
as well as oscillations in supernova envelopes.
Where neutrino densities are relatively high, however,
as in neutron star mergers, the problem must be solved globally.
To date, no method has been devised to handle this problem in systems lacking
spherical symmetry.

However, we can solve a similar but tractable problem along a single ray.
We assume that all the neutrino rays intersecting an event along a given test ray
have undergone the same flavor evolution history as that of the test ray:
i.e.\ the evolution matrix $S$ is the same for all rays sharing that event.
This is the so-called single-angle approximation,
which is widely used in the supernova oscillation literature
and has been shown to be qualitatively faithful in those environments.
We note, however, that recent studies have discovered 1) spherically symmetric
configurations for which a single-angle calculation produces qualitatively
different flavor evolution behavior than a full multi-angle calculation
\cite{vlas2018-multiangle},
and 2) azimuthally-symmetric configurations for which the single-angle
approximation masks certain instabilities
in the flavor evolution \cite{wu2017-fast_neutrino_conversions}.

The formalism of the single-angle approximation requires knowledge of
the unoscillated neutrino contribution to the Hamiltonian matrix
along a given test trajectory.
This is a function of the unoscillated neutrino self-interaction potential,
which for the $\alpha$-th flavor is:
\begin{equation}
  \label{eqn:V_nu_alpha}
  V_{\nu_\alpha,0}(\varepsilon,\ell_\mu) =
  \frac{\sqrt{2}G_{\rm F}}{(2\pi)^3} \varepsilon^2
  \oint d\Omega'(1-\varpi')f^{\nu_\alpha}(\varepsilon,\ell'_\gamma),
\end{equation}
with the test ray propagating in direction $\ell_\mu$,
ambient rays propagating in directions $\ell'_\gamma$,
and the cosine of the angle between these given by
$\varpi'=\psi^{\mu\gamma} \ell_\mu \ell'_\gamma$.
As in the moment equations (Eqns.~\ref{eqn:J}-\ref{eqn:Ka}),
the integral is taken over all directions $\ell'_\gamma$.

@@

If conditions are right, a resonant flavor transition
introducing significant mixing may occur very near
the point where neutrinos begin free-streaming.
The matter-neutrino resonance
\cite{malk2012-mnr_1,malk2015-mnr_2,malk2016-mnr_3},
can occur where the matter potential and
neutrino self-interaction potentials cancel,
or where $V_e + V_{\nu\nu,0}=0$, with
\begin{align}
  \label{eqn:V_e}
  V_e &= \sqrt{2} G_{\rm F} \frac{\varrho}{m_N} Y_e, \\
  \label{eqn:V_nunu0}
  V_{\nu\nu,0} &= \int d\varepsilon
  \big(V_{\nu_e,0}(\varepsilon)-V_{\bar{\nu}_e,0}(\varepsilon)\big),
\end{align}
with $\varrho$ the rest density and $m_{N}$ the nucleon mass.
The matter potential $V_e$ is always positive;
and far outside the accretion disks formed in neutron star mergers,
in which the disrupted neutron-star matter is rapidly releptonizing,
the total unoscillated neutrino self-interaction potential $V_{\nu\nu,0}$
is large and negative.

We examine this effect in the post-merger configuration already analyzed
in Sec.~\ref{ssec:test_disk_comparison}.
We calculate the self-interaction potential along a radial coordinate trajectory
making an angle $\theta=25^\circ$ with the rotation axis,
at $N_r=7$ positions $r\in\{30,50,82,135,223,368,608\}$~km.
We sample distribution functions at each of these positions over a grid
with extents $N_\varepsilon=30$, $N_A=200$, $N_B=30$;
energies range over $\varepsilon\in(0,100)$~MeV;
polar angles range over $\cos A \in ((\cos A)_{\rm min}, 1)$, with
$(\cos A)_{\rm min}\in\{-1,-1,-1,0.488,0.849,0.947,0.981\}$
for each of the $N_r$ positions;
and azimuthal angles range over the whole sky $B\in[0,2\pi)$.
We interpolate the logarithm of the self-interaction potentials
(Eqn.~\ref{eqn:V_nu_alpha})
in path length $\log r$ along the ray,
by fitting a 3rd order spline with continuous derivative,
across the $N_r$ observation points.
For $r>608$~km we extrapolate the self-interaction potentials using the
geometric fall-off of $r^{-4}$ applicable to the far-field limit of the
self-interaction potential \cite{malk2016-mnr_3}.

@@

In fact a spherical steady-state wind model, though providing an adequate
backdrop for the qualitative study presented in this section,
is less than ideal for this post-merger configuration,
most obviously because in the 10~ms since merger,
ejecta with the greatest velocities around $0.3c$
will have reached no further than $r_{\rm max} \sim 10^8$~cm.
Additionally, the true radial profile of the ejecta from this merger
(which our computational model does not follow)
will have many more features inside this radius, including shock jumps.
However, this model density profile is adequate as a backdrop to our study here
since we expect the remnant to present similar neutrino emission over a thermal
timescale of a few tens of milliseconds
while the matter field propagates out to larger radii.

@@

We also solve for the flavor evolution of this system,
integrating $S$ according to Eqn.~\ref{eqn:schroedinger},
as described in \cite{zhu2016-mnr_nsns_remnant}.
We show the survival probabilities for $\nu_e$ and $\bar{\nu}_e$
for the \emph{noscat} treatment in Fig.~\ref{fig:survival-noscat},
and for the \emph{scat} treatment in Fig.~\ref{fig:survival-scat}.
The survival probability is the probability that a neutrino, if measured,
will be found to be in its original flavor state.
The survival probabilities are computed at each point along the trajectory
from the absolute square of the diagonal terms of the flavor evolution matrix,
$P_{\nu_\alpha \to \nu_\alpha} = |S_{\alpha \alpha}|^2$.
When $P_{\nu_e \to \nu_e}$ decreases,
as can be seen for example in Fig.~\ref{fig:survival-scat} for $r>400$~km,
some of the electron-type neutrinos have oscillated into
mu- or tau-type neutrinos.

@@

\begin{figure}
  \resizebox{\columnwidth}{!}{\input{fig-survival_prob-noscat_ih.tex}}
  % Flavor evolution by Yonglin, see emails 5.9.18, 5.16.18
  % Data: $DIR/300-1499kmACC12Step2NE30IH051418m12sc.dat2p
  % Script: $DIR/plot-noscat.gp
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170412-self_interaction_potentials_M12_M12-LS220-M1/Run03/RunBA/RunBA-025deg-noscat/flavor_evolution
  \caption{Neutrino survival probabilities for the \emph{noscat} case.
    In the lower panel we reproduce the self-interaction potential $V_{\nu\nu,0}$
    and matter potential $V_e$ from Fig.~\ref{fig:V_nunu-noscat}.
    In this case, the neutrinos undergo collective neutrino oscillation
    beginning around $r\sim700$~km,
    with both the electron neutrinos and antineutrinos almost converting to mu
    or tau neutrinos completely.
  }
  \label{fig:survival-noscat}
\end{figure}

\begin{figure}
  \resizebox{\columnwidth}{!}{\input{fig-survival_prob-scat_ih.tex}}
  % Flavor evolution by Yonglin, see emails 5.9.18, 5.16.18
  % Data: $DIR/300-1499kmACC12Step2NE30IH051418m12sc.dat2p
  % Script: $DIR/plot-scat.gp
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170412-self_interaction_potentials_M12_M12-LS220-M1/Run03/RunBA/RunBA-025deg-scat/flavor_evolution
  \caption{Same as Fig.~\ref{fig:survival-noscat},
    but for the \emph{scat} case.
    In this case the neutrinos undergo a standard MNR transition
    beginning around $r\sim400$~km,
    with the electron neutrinos converting to mu or tau neutrinos almost
    completely,
    and the electron antineutrinos oscillating back to their original flavor
    after partially converting to mu and tau antineutrinos.
  }
  \label{fig:survival-scat}
\end{figure}

@@

In addition to the fluid velocity and Lorentz factor, described above, the
other fluid state variables we use from our hydrodynamic simulations
are rest density $\varrho=m_b n_b$, temperature $T$, and electron fraction
\begin{equation}
  Y_e = \frac{n_{e^-} - n_{e^+}}{n_b},
\end{equation}
where $n_{e^-}$, $n_{e^+}$, and $n_b$ are the number densities of electrons,
positrons, and baryons, and $m_b$ is the average baryon mass.
